Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-11-29 SEIR.tex
Updates from Overleaf
Figures and edits

This chapter uses mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving model's state and parameters have a biological and/or epidemiological interpretation.
Of particular interest is that the basic reproduction number, $R_0$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
However, they make stronger assumptions about the disease and population behaviour than statistical models.

@@

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item For any short period of time when $s(t)$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth~\autocite[section 1.2]{diekmannMathematical}.
    A notable example is the early part of an epidemic when $s(t) \approx 1$.
    This is the phase of initial exponential growth of an epidemic, which produces the characteristic early epidemic curves.
    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
    \item The basic reproduction number for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
    Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = R_0 s(t)$~\autocite{pellisEstimation}.
    This is the basic reproduction number reduced because some infections are avoided due to individuals being immune.
    \item The population is closed.
    That is, the total population size is constant, and no infections are imported from outside the population.
    Births and deaths of individuals are also excluded, although they can easily be incorporated in situations where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
    \item In addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
    \item Immunity is assumed to last until the end of the modelled period~\autocite[61]{andersonInfectious}.
    In many contexts, including here, this assumption is justified because the modelled period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For modelling longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
\end{itemize}

@@

The reproduction number in terms of the parameters is unchanged by this extension, \ie $R_0 = \beta / \gamma$.

@@

The reproduction number in terms of the parameters is unchanged by this extension, \ie $R_0 = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.

@@

Reproduction numbers in models with structured populations need to be careful to consider the distribution of infections across the strata.
The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata; this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$. 
However, how to average these contributions to calculate the basic reproduction number is not obvious.
The standard solution to this was proposed by \textcite{diekmannDefinition} (\textcite[chapter 7]{diekmannMathematical} provides a more gentle, but still rigorous, explanation).
First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{0}$) relies only on $\matr{\beta}$.
This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
The basic reproduction number is then defined as $R_0 = D(\matr{\beta}) / \gamma$.
This $R_0$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
The important threshold that exponential growth occurs (in expectation) iff $R_0 > 1$ remains.

@@

From the system of ODEs\todo{might need to reorder this because I've not presented these yet.. } and the knowledge that all the compartments are growing exponentially at rate $\psi$, one can derive the following for the initial state of the compartments:
\begin{align}
    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \kappa_0} \\
    \vec{p_1} &= \vec{p_0} \frac{\kappa_0}{\psi + \kappa_Y} \\
    \vec{p_2} &= \vec{p_1} \frac{\kappa_Y}{\psi + \kappa_{X1}} \\
    \vec{p_3} &= \vec{p_2} \frac{p_x \kappa_{X1}}{\psi + \kappa_{X2}}.
\end{align}

@@

All fixed values, including $\matr{\kappa}$, used the values from the main analysis.
This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
In particular, the October half-term school holiday is visible, especially in the school-age groups and the November lockdown is visible to some extent in all age groups.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/sim/data}}
    \caption[Simulated data]{%
        Incidence (top) and prevalence (bottom) for the 100 simulated data sets as a proportion of each stratum.
        Prevalence shown excludes 
    }
    \label{SEIR:fig:sim-data}
\end{figure}

@@

To estimate the posterior, I ran a single MCMC chain per simulated data set for 500,000 iterations, discarding the first 300,000 iterations as burn-in, and thinning every 100 iterations.
Therefore, posterior calculations are based on 2,000 samples per data set.
On some occasions, the proposal distribution's covariance matrix became singular (or close to singular).
This means it is not possible to propose a new value.
In these cases, I restarted the chain with a new seed, discarding all information from the original chain; this was required at most once per chain.

@@

Unlike most prior work, this modelling is based on a data source that is representative of the general population.
Therefore, the results are have fewer biases and informed by data to a greater extent than this work.
There is little-to-no ascertainment bias in the CIS, unlike sources relying on testing behaviour (in the UK, pillar 2 testing).
The CIS allows data-informed estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults\todo{cite something about mildness in children or age distribution of severe events}.

The parameters $\beta_t$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
\todo{discuss the results and implications for policy or understanding transmission}
Many factors contribute to the pattern in the $\beta_t$ values, which have a minima near the start of December.
While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
The decreases in $\beta_t$ over the autumn could be due to increased caution as cases and deaths increased (\eg masking or stricter adherence to social distancing),\todo{cite something about caution and reaction to rising cases, or survey data showing this happened - CoMix?} or the impact of local measures in the highest transmission areas\todo{cite local measures}.
The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
A multiple variant model might be able to capture this rise more accurately.
Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
Then, each variant would be explicitly included in the model to fit to this data.

@@

>>>>

This chapter uses mechanistic models for estimating the transmission of SARS-CoV-2, an alternative to the statistical approach in \cref{E-backcalc}.
Mechanistic models explicitly represent the infection process driving the spread of infectious disease, giving model's state and parameters have a biological and/or epidemiological interpretation.
Of particular interest is that the basic reproduction number, $\R$, and the effective reproduction number, $R_e(t)$, which can both be calculated as functions of the model parameters; using \cref{E-backcalc}'s approach, additional modelling would be required to estimate these numbers.
Furthermore, mechanistic models can be used for scenario-based modelling, which simulates the effect of interventions.
All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
However, they make stronger assumptions about the disease and population behaviour than statistical models.

@@

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item For any short period of time when $s(t)$ is approximately constant, the system of ODEs can be solved analytically to produce exponential growth~\autocite[section 1.2]{diekmannMathematical}.
    A notable example is the early part of an epidemic when $s(t) \approx 1$.
    This is the phase of initial exponential growth of an epidemic, which produces the characteristic early epidemic curves.
    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
    \item The basic reproduction number for this model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
    Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
    \item The effective reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
    This is the basic reproduction number reduced because some infections are avoided due to individuals being immune.
    \item The population is closed.
    That is, the total population size is constant, and no infections are imported from outside the population.
    Births and deaths of individuals are also excluded, although they can easily be incorporated in situations where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
    \item In addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
    \item Immunity is assumed to last until the end of the modelled period~\autocite[61]{andersonInfectious}.
    In many contexts, including here, this assumption is justified because the modelled period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For modelling longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
\end{itemize}

@@

The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$.

@@

The reproduction number in terms of the parameters is unchanged by this extension, \ie $\R = \beta / \gamma$, however, the growth rate in the exponential phase will be slower than before~\autocite[41]{keelingModeling}.

@@

Reproduction numbers in models with structured populations need to be careful to consider the distribution of infections across the strata.
The WAIFM allows calculation of the number of secondary infections generated by a single infection in the $a$th strata; this follows directly from its definition as $\frac{1}{\gamma} \sum_{a' \in \set{A}} \beta_{aa'}$. 
However, how to average these contributions to calculate the basic reproduction number is not obvious.
The standard solution to this was proposed by \textcite{diekmannDefinition} (\textcite[chapter 7]{diekmannMathematical} provides a more gentle, but still rigorous, explanation).
First, they show that the distribution of infections across the strata during the initial period of exponential growth (when $\vec{s} \approx \vec{1}$) relies only on $\matr{\beta}$.
This distribution is proportional to the \emph{dominant eigenvector} (the eigenvector corresponding to the largest eigenvalue) of $\matr{\beta}$.
Denote by $D(\matr{\beta})$ the \emph{dominant eigenvalue} (the largest eigenvalue).
The basic reproduction number is then defined as $\R = D(\matr{\beta}) / \gamma$.
This $\R$ keeps the same interpretation as in the homogeneous case, \ie the average number of secondary infections caused by a single infection in a fully susceptible population, where the average is taken over the distribution implied by the dominant eigenvector.
The important threshold that exponential growth occurs iff $\R > 1$ remains.

@@

From the system of ODEs and the knowledge that all the compartments are growing exponentially at rate $\psi$, one can derive the following for the initial state of the compartments:
\begin{align}
    \vec{i_1} &= \vec{i_0} \left(1 + \frac{2\gamma}{\psi + 2\gamma} \right)^{-1} \\
    \vec{i_2} &= \vec{i_0} - \vec{i_1} \\
    \vec{e_2} &= \vec{i_1} \frac{\psi + 2\gamma}{2\sigma} \\
    \vec{e_1} &= \vec{e_2} \frac{\psi + 2\sigma}{2\sigma} \\
    \vec{s} &= \vec{\pi} - \vec{i_0} - \vec{e_1} - \vec{e_2} \\
    \vec{p_0} &= \vec{e_1} \frac{\psi + 2\sigma}{\psi + \kappa_0} \\
    \vec{p_1} &= \vec{p_0} \frac{\kappa_0}{\psi + \kappa_Y} \\
    \vec{p_2} &= \vec{p_1} \frac{\kappa_Y}{\psi + \kappa_{X1}} \\
    \vec{p_3} &= \vec{p_2} \frac{p_x \kappa_{X1}}{\psi + \kappa_{X2}}.
\end{align}

@@

All fixed values, including $\matr{\kappa}$, used the values from the main analysis.
This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
In particular, the October half-term school holiday is visible, especially in the school-age groups and the November lockdown is visible to some extent in all age groups.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/sim/data}}
    \caption[Simulated data]{%
        Incidence (top) and prevalence (bottom) for the 100 simulated data sets as a proportion of each stratum.
        Prevalence shown is $\bar{p}_{kat}$, the true population prevalence on each day, excluding sampling noise.
        Each line represents one simulation.
    }
    \label{SEIR:fig:sim-data}
\end{figure}

@@

To estimate the posterior, I ran a single MCMC chain per simulated data set for 500,000 iterations, discarding the first 300,000 iterations as burn-in, and thinning every 100 iterations.
Therefore, posterior calculations are based on 2,000 samples per data set.
On some occasions, the proposal distribution's covariance matrix became singular (or close to singular).
\todo{Update with fitting issues if they reoccur on final runs}
This means it is not possible to propose a new value.
In these cases, I restarted the chain with a new seed, discarding all information from the original chain; this was required at most once per chain.

To check that the inference was working correctly, I calculated the coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter and simulation.
I define coverage as the proportion of simulations where the true value of the parameter in that simulation was within the credible interval.
The coverage for each parameter should equal the nominal probability (\eg the coverage should be 50\% for the 50\% credible intervals).
I constructed a confidence interval for the coverage using the Pearson-Klopper method implemented in the binom package~\autocite{binom1-1}.

Overall, coverage was\todo{comment on sim study coverage}
\begin{figure}
    \includegraphics{SEIR/sim/coverage}
    \caption[Coverage of simulation study]{%
        Coverage of the 50\%, 75\%, and 95\% central credible intervals for each parameter.
        The lines shows the nominal coverage.
    }
    \label{SEIR:fig:sim-coverage}
\end{figure}

@@

Unlike most prior work, this modelling is based on a data source that is representative of the general population.
Therefore, the results are have fewer biases and informed by data to a greater extent than this work.
There is little-to-no ascertainment bias in the CIS, unlike sources relying on testing behaviour (in the UK, pillar 2 testing).
The CIS allows data-informed estimates of transmission in children, unlike sources relying on severe events that occur almost entirely in adults~\autocite{bhopalChildren}.

The parameters $\beta_t$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
\todo{discuss the results and implications for policy or understanding transmission}
Many factors contribute to the pattern in the $\beta_t$ values, which have a minima near the start of December.
While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
The decreases in $\beta_t$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas\todo{cite local measures}.
The increases in $\beta_t$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
A multiple variant model might be able to capture this rise more accurately.
Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
Then, each variant would be explicitly included in the model to fit to this data.

@@

\todo[inline]{write conclusion and add to discussion once comparison with backcalculation is made}
